Computational Modeling HW 1

### 2015-11-02 classical/latex/a4.tex

We want to find the velocity at which $m\dot{y}$ is maximized. We find when the derivative of this with respect to time is zero, plug in for the values we find and go from there:

\begin{align*}
  \frac{\partial}{\partial t} m\dot{y} &= \dot{m}\dot{y} + m\ddot{y} = 0\\
  m\ddot{y} &= -mg - \dot{m}v_{ex}\\
  \dot{m}\dot{y} &= mg + \dot{m}v_{ex}\\
  \dot{y} &= \frac{mg + \dot{m}v_{ex}}{\dot{m}}\\
\end{align*}

@@

\begin{align*}
  m\ddot{y} &= -\frac{b}{m}\dot{y} - \dot{m}v_{ex}\\
\end{align*}

@@

\begin{align*}
  SA &= \int_{y_1}^{y_2} dA = \int_{y_1}^{y_2} 2\pi\sqrt{x^2+y^2}\sqrt{x'^2+1}dy\\
  f(x, x', y) &= 2\pi\sqrt{x^2+y^2}\sqrt{x'^2+1}\\
\end{align*}

\begin{align*}
  \frac{\partial f}{\partial x} &- \frac{\partial}{\partial y} \frac{\partial f}{\partial x'} = 0\\
  \frac{\partial f}{\partial x} &= \frac{2\pi x\sqrt{x'^2+1}}{\sqrt{x^2+y^2}}\\
  \frac{\partial}{\partial y} \frac{\partial f}{\partial x'} &= 0\\
\end{align*}

So...
\begin{align*}
  \frac{2\pi x\sqrt{x'^2+1}}{\sqrt{x^2+y^2}} = 0\\
\end{align*}

@@

\begin{align*}
  T &= \int_{\theta_1}^{\theta_2} \frac{1}{v}ds = \int_{\theta_1}^{\theta_2} \frac{a}{cr^2}\sqrt{r^2+r'^2}d\theta\\
  f(r,r',\theta) &= \frac{a}{cr^2}\sqrt{r^2+r'^2}d\theta\\
\end{align*}

\begin{align*}
  \frac{\partial f}{\partial r} - \frac{\partial}{\partial \theta} \frac{\partial f}{\partial r'} &= 0\\
  \frac{\partial}{\partial \theta} \frac{\partial f}{\partial r'} &= 0\\
\end{align*}

And so:

\begin{align*}
    \frac{\partial f}{\partial r} = 0\\
\end{align*}

Meaning

\end{document}

>>>>

\begin{align*}
  m\ddot{y} &= -\dot{m}V_{ex}\\
  m\frac{dv}{dt} &= -\frac{dm}{dt}V_{ex}\\
  dv &= -\frac{dm}{m}V_{ex}\\
  \int_0^v dv &= \int_{m_0}^m-\frac{dm}{m}V_{ex}\\
  v &= -\log(\frac{m}{m_0})V_{ex}\\
  m &= m_0e^{\frac{-v}{V_{ex}}}\\
\end{align*}

We want to find the velocity which minimizes momentum...

\begin{align*}
  p &= mv\\
  \frac{dp}{dv} &= m + v\frac{dm}{dv} = 0\\
  m + \frac{m_0v}{V_{ex}}e^{\frac{-v}{V_{ex}}} &= 0\\
  m &= \frac{m_0v}{V_{ex}}e^{\log(\frac{m}{m_0})}\\
  v &= V_{ex}\\
\end{align*}

This problem has an interesting result. It isn't intuitive to me why it is the case - it probably has something to do with momentum conservation. The derivation itself was fairly simple, just another application of separation of variables. I like getting practice using this and other methods to solve differential equations.\\

@@

\begin{align*}
  (m_0+\dot{m}t)\ddot{y} &= -b\dot{y} - \dot{m}v_{ex}\\
  \frac{d\dot{y}}{dt} &= \frac{-b\dot{y} - \dot{m}v_{ex}}{(m_0+\dot{m}t)}\\
  \int_0^v \frac{d\dot{y}}{-b\dot{y} - \dot{m}v_{ex}} &= \int_0^t \frac{1}{(m_0+\dot{m}t)}dt\\
  \frac{-1}{b}\log(\frac{bv + \dot{m}v_{ex}}{\dot{m}v_{ex}}) &= \frac{1}{\dot{m}}\log(\frac{m_0+\dot{m}t}{m_0})\\
  -\dot{m}\log(\frac{bv + \dot{m}v_{ex}}{\dot{m}v_{ex}}) &= b\log(\frac{m_0+\dot{m}t}{m_0})\\
  \frac{bv + \dot{m}v_{ex}}{\dot{m}v_{ex}} &= (\frac{m_0+\dot{m}t}{m_0})^{\frac{b}{\dot{m}}}\\
  v &= \frac{1}{b}\Big(\dot{m}v_{ex}(\frac{m_0+\dot{m}t}{m_0})^{\frac{-b}{\dot{m}}} - \dot{m}v_{ex}\Big)\\
  v &= \frac{kV_{ex}}{b}\Big(1 - (\frac{m}{m_0})^{\frac{b}{k}} \Big)\\
\end{align*}

This was another interesting problem. I think I'm getting really experienced at using separation of variables to solve equations. The equation for velocity makes sense - as the $m/m_0$ fraction decreases, the velocity increases. Since the fraction m/m0 is always less than one, the velocity is always possible. It is interesting that the terminal velocity once all the mass has been expended (if all the rocket is actually fuel) is $kV_{ex}/b$ - this is good to know.\\

@@

\begin{align*}
  SA &= \int_{y_1}^{y_2} dA = \int_{y_1}^{y_2} 2\pi y\sqrt{y'^2+1}dx\\
  f(y, y', x) &= 2\pi y\sqrt{y'^2+1}\\
\end{align*}

\begin{align*}
  f - y'\frac{\partial f}{\partial y'} &= C\\
  y\sqrt{y'^2+1} - \frac{yy'^2}{\sqrt{y'^2+1}} &= C\\
  y\sqrt{y'^2+1} &= C + \frac{yy'^2}{\sqrt{y'^2+1}}\\
  y(y'^2+1) &= C\sqrt{y'^2+1} + yy'^2\\
  y &= C\sqrt{y'^2+1}\\
  \frac{y^2}{C^2} - 1 &= y'^2\\
  y^2 - C^2 &= y'^2C^2\\
  \frac{\sqrt{y^2 - C^2}}{C} &= y'\\
  \frac{\sqrt{y^2 - C^2}}{C} &= y'\\
  \frac{dx}{dy} = \frac{1}{y'} &= \frac{C}{\sqrt{y^2 - C^2}}\\
  x = \int \frac{dx}{dy}dy &= \int \frac{1}{y'} = \frac{C}{\sqrt{y^2 - C^2}}dy\\
  x = C\cosh^{-1}(\frac{y}{C})+b\\
  y = C\cosh(\frac{x-b}{C})\\
\end{align*}

So we can see that a hyperbolic cosine describes the smallest surface which connects two points. We can find the values of b and C by solving our result for the initial and final points, which must be done with a computer. This corresponds to a surface which starts at one point, bows inwards, then arcs outward again. This problem was a cool introduction to the power of using the Euler-Lagrange equation to find surfaces. It was actually fairly easy to get the solution, given the difficulty of solving the problem by any other means.\\

@@

\begin{align*}
  T &= \int_{\theta_1}^{\theta_2} \frac{1}{v}ds = \int_{\theta_1}^{\theta_2} \frac{a}{cr^2}\sqrt{r^2+r'^2}d\theta\\
  f(r,r',\theta) &= \frac{a}{cr^2}\sqrt{r^2+r'^2}\\
\end{align*}

\begin{align*}
  f - r'\frac{\partial f}{\partial r'} &= C\\
  \frac{a}{cr^2}\sqrt{r^2+r'^2} - \frac{ar'^2}{cr^2\sqrt{r^2+r'^2}} &= C\\
  \frac{1}{r^2}\sqrt{r^2+r'^2} - \frac{r'^2}{r^2\sqrt{r^2+r'^2}} &= \frac{Cc}{a}\\
  \sqrt{r^2+r'^2} &= \frac{Ccr^2}{a} + \frac{r'^2}{\sqrt{r^2+r'^2}}\\
  r^2+r'^2 &= \frac{Ccr^2}{a}\sqrt{r^2+r'^2} + r'^2\\
  \frac{a^2}{C^2c^2} &= r^2+r'^2\\
  \sqrt{\frac{a^2}{C^2c^2} - r^2} &= \frac{dr}{d\theta}\\
  \int d\theta &= \int \frac{dr}{\sqrt{\frac{a^2}{C^2c^2} - r^2}}\\
  \theta &= \tan^{-1}\Big( \frac{c^2r\sqrt(\frac{a^2C^2-c^2r^2}{c^2})}{-a^2C^2+c^2r^2}   \Big)\\
  \tan(\theta) &= \frac{c^2r\sqrt{\frac{a^2C^2-c^2r^2}{c^2}}}{-a^2C^2+c^2r^2}\\
  r(\theta) &= \frac{C\tan(\theta)}{\sqrt{1+\tan(\theta)^2}}
\end{align*}

This was a very interesting problem. We arrived at a concise solution for a very complicated physical problem using relatively straightforward techniques. It was interesting to note that the actual value of the physical constants is not important, since the path is minimized this particular way for any value of a and c. The C in our final expression is just a constant multiple that modifies the radius of the circle.
\end{document}

### 2015-11-02 comp_model/latex/a2.tex

\section{1.4: Trapezoidal method coefficients}
\begin{equation}
  \sum_{j=0}^r \alpha_jV_{n+j} = h \sum_{j=0}^r \beta_j f(t_{n+j}, V_{n+j})\\
\end{equation}

The trapezoidal method is this:\\
\begin{equation}
  V_{n+1} = V_n + h\Big( \frac{1}{2}f(V_n, t_n) + \frac{1}{2}f(V_{n+1}, t_{n+1}) \Big)
\end{equation}

By examining these two equations, we can see that:

\begin{align}
  \alpha_0 = -1 &\hspace{2cm} \beta_0 = \frac{1}{2}\\
  \alpha_1 = 0 &\hspace{2cm} \beta_1 = 1\\
  \alpha_2 = 0 &\hspace{2cm} \beta_2 = 1\\
  ...  &\hspace{2cm} ...\\
  \alpha_{r-1} = 0 &\hspace{2cm} \beta_{r-1} = 1\\
  \alpha_r = 1 &\hspace{2cm} \beta_1 = \frac{1}{2}\\    
\end{align}

\begin{equation}
  V_{n+1} = V_n + h\Big( \frac{1}{2}f(V_n, t_n) + \frac{1}{2}f(V_{n+1}, t_{n+1}) \Big)
\end{equation}

If we let our function f take the following form: $f(x, t) = \alpha x$ where $x(0) = x_0$, we can see that:

\begin{align}
  V_{n+1} &= V_n + \frac{h}{2}\Big( f(V_n, t_n) + f(V_{n+1}, t_{n+1}) \Big)\\
  V_{n+1} &= V_n + \frac{h}{2}\Big( \alpha V_n + \alpha V_{n+1} \Big)\\
  V_{n+1} &= V_n + \frac{\alpha hV_n}{2} + \frac{\alpha hV_{n+1}}{2}\\
  V_{n+1} \Big(1 - \frac{\alpha h}{2}\Big) &= V_n \Big(1 + \frac{\alpha h}{2}\Big) \\
  V_{n+1} &= V_n \frac{1 + \frac{\alpha h}{2}}{1 - \frac{\alpha h}{2}} \\
  V_{n+1} &= V_n \frac{2 + \alpha h}{2 - \alpha h} \\
  V_{n+1} &= V_0 \Big(\frac{2 + \alpha h}{2 - \alpha h}\Big)^{n+1} \\    
\end{align}

From this we can see that as long as the following condition is satisfied, we will have a converging solution:

\begin{align}
  |2 + \alpha h| \leq |2 - \alpha h|\\
\end{align}
\section{1.9: Order of Accuracy for Trapezoidal and Euler Methods}
We want to analytically represent the error terms for the Euler and Trapezoidal methods and show how they depend on our timestep, $h$.\\

First we solve for the behavior of our function using a Taylor Series. Let $x(t)$ be our function and $v(t) = \frac{dx(t)}{dt}$. Let $N = \frac{t}{h}$, where t is the current time and h the timestep. Finally, let $x(t_0) = 0$.\\

\begin{align}
  x(t) &= \sum_{i=0}^{i=N-1} \int_{t_i}^{t_{i+1}} x'(t)dt = \sum_{i=0}^{i=N-1} \int_{t_i}^{t_{i+1}} v(t)dt\\
  v(t) &= v(t_i) + (t-t_i)v'(t_i) + (t-t_i)^2v''(t_i) + ...\\
  \int_{t_i}^{t_{i+1}} v(t)dt &\approx \int_{t_i}^{t_{i+1}} v(t_i) + (t-t_i)v'(t_i) + \frac{1}{2}(t-t_i)^2v''(t_i)  dt\\
\end{align}

We now do a u-substitution, where $S = \frac{t - t_i}{t_{i+1} - t_i}$ and $h = t_{i+1} - t_i$:

\begin{align}
  \int_{t_i}^{t_{i+1}} v(t)dt &\approx h\int_i^1 v(t_i) + hSv'(t_i) + \frac{1}{2}h^2S^2v''(t_i) dS\\
  \int_{t_i}^{t_{i+1}} v(t)dt &\approx h \big( v(t_i)S + \frac{hS^2}{2}v'(t_i) + \frac{h^2S^3}{6}v''(t_i) \big) |_0^1 \\
  \int_{t_i}^{t_{i+1}} v(t)dt &\approx h \big( v(t_i) + \frac{h}{2}v'(t_i) + \frac{h^2}{6}v''(t_i) \big) \\
  \int_{t_i}^{t_{i+1}} v(t)dt &\approx hv(t_i) + \frac{h^2}{2}v'(t_i) + \frac{h^3}{6}v''(t_i) \\
\end{align}

This gives us a fairly exact value for $\int_{t_i}^{t_{i+1}} v(t)dt$. We now need to compare this with the approximations that our two methods make. For Euler's method, $\int_{t_i}^{t_{i+1}} v(t)dt = hf(t_i)$. For the Trapezoidal method, $\int_{t_i}^{t_{i+1}} v(t)dt = \frac{h}{2}\Big( f(t_{i+1}) + f(t_{i}) \Big)$. We will first plug in the Euler's method approximation to the above expression to find its error term.

\begin{align}
  \int_{t_i}^{t_{i+1}} v(t)dt &= hv(t_i)\\
  \int_{t_i}^{t_{i+1}} v(t)dt &= hv(t_i) + \frac{h^2}{2}v'(t_i) + \frac{h^3}{6}v''(t_i)\\
  hv(t_i) &= hv(t_i) + \frac{h^2}{2}v'(t_i) + \frac{h^3}{6}v''(t_i)\\
  -\frac{h}{3}v''(t_i) &= v'(t_i)\\  
\end{align}

We now substitute this back into our original result:

\begin{align}
  \int_{t_i}^{t_{i+1}} v(t)dt &= hv(t_i) + \frac{h^2}{2}v'(t_i) + \frac{h^3}{6}v''(t_i) \\
  \int_{t_i}^{t_{i+1}} v(t)dt &= hv(t_i) + \frac{h^2}{2}(-\frac{h}{3}v''(t_i)) + \frac{h^3}{6}v''(t_i) \\
  \int_{t_i}^{t_{i+1}} v(t)dt &= hv(t_i) + -\frac{h^3}{4}(v''(t_i)) + \frac{h^3}{6}v''(t_i) \\
  \int_{t_i}^{t_{i+1}} v(t)dt &= hv(t_i) + -\frac{h^3}{12}(v''(t_i))\\
\end{align}

As we can see, the error for the forward Euler's method is $-\frac{h^3}{12}(v''(t_i))$ per timestep. We now plug this result into our summed equation to find the total error for all timesteps:

\begin{align}
  x(t) &= \sum_{i=0}^{i=N-1} \int_{t_i}^{t_{i+1}} x'(t)dt = \sum_{i=0}^{i=N-1} \int_{t_i}^{t_{i+1}} v(t)dt\\
  x(t) &= \sum_{i=0}^{i=N-1} \int_{t_i}^{t_{i+1}} x'(t)dt = \sum_{i=0}^{i=N-1} hv(t_i) + -\frac{h^3}{12}(v''(t_i))\\
  x(t) &= \sum_{i=0}^{i=N-1} \int_{t_i}^{t_{i+1}} x'(t)dt = \sum_{i=0}^{i=N-1} hv(t_i) + -\frac{h^3}{12}(v''(t_i))\\  
\end{align}

We only care about the error term, so we will focus on that. We substitute $N$ with $\frac{t - t_0}{h}$ and replace $v''(t_i)$ with a term that represents the average value of $v''(t_i)$ over the whole sum: $v''_{ave}$.

\begin{align}
  E &= \sum_{i=0}^{i=N-1} -\frac{h^3}{12}(v''(t_i))\\
  E &= -N\frac{h^3}{12}v''_{ave}\\
  E &= -\frac{t - t_0}{h} \frac{h^3}{12}v''_{ave}\\
  E &= -(t - t_0)\frac{h^2}{12}v''_{ave}\\    
\end{align}

We do the same thing for the Trapezoidal method. We first find the error term:

\begin{align}
  \int_{t_i}^{t_{i+1}} v(t)dt &= hv(t_i)\\
  \int_{t_i}^{t_{i+1}} v(t)dt &= hv(t_i) + \frac{h^2}{2}v'(t_i) + \frac{h^3}{4}v''(t_i)\\
  \frac{h}{2}\Big(v(t_{i+1}) + v(t_i)\Big) &= hv(t_i) + \frac{h^2}{2}v'(t_i) + \frac{h^3}{4}v''(t_i)\\
  \frac{h}{2}\Big(-v(t_{i+1}) + v(t_i)\Big) - \frac{h^3}{4}v''(t_i) &= \frac{h^2}{2}v'(t_i)\\
  \frac{1}{h}\Big(-v(t_{i+1}) + v(t_i)\Big) - \frac{h}{2}v''(t_i) &= v'(t_i)\\
\end{align}

We then substitute this back into our original result:

\begin{align}
  \int_{t_i}^{t_{i+1}} v(t)dt &= hv(t_i) + \frac{h^2}{2}v'(t_i) + \frac{h^3}{6}v''(t_i) \\
  \int_{t_i}^{t_{i+1}} v(t)dt &= hv(t_i) + \frac{h^2}{2}\Big(\frac{1}{h}\Big(v(t_{i+1}) - v(t_i)\Big) - \frac{h}{4}v''(t_i) \Big) + \frac{h^3}{6}v''(t_i) \\
  \int_{t_i}^{t_{i+1}} v(t)dt &= \frac{h}{2}\big(v(t_{i+1}) + v(t_i)\big) - \frac{h^3}{4}v''(t_i) + \frac{h^3}{6}v''(t_i) \\
  \int_{t_i}^{t_{i+1}} v(t)dt &= \frac{h}{2}\big(v(t_{i+1}) + v(t_i)\big) - \frac{h^3}{12}v''(t_i)\\    
\end{align}

From this we can see that the error term for the trapezoidal rule is $-\frac{h^3}{12}v''(t_i)$ per timestep. We then sum over all timesteps as above:

\begin{align}
    x(t) &= \sum_{i=0}^{i=N-1} \int_{t_i}^{t_{i+1}} x'(t)dt = \sum_{i=0}^{i=N-1} \int_{t_i}^{t_{i+1}} v(t)dt\\
  x(t) &= \sum_{i=0}^{i=N-1} \int_{t_i}^{t_{i+1}} x'(t)dt = \sum_{i=0}^{i=N-1} \frac{h}{2}\big(v(t_{i+1}) + v(t_i)\big) - \frac{h^3}{12}v''(t_i)\\
  x(t) &= \sum_{i=0}^{i=N-1} \int_{t_i}^{t_{i+1}} x'(t)dt = \sum_{i=0}^{i=N-1} \frac{h}{2}\big(v(t_{i+1}) + v(t_i)\big) + \sum_{i=0}^{i=N-1} -\frac{h^3}{12}v''(t_i)\\
\end{align}

We finally solve for the error:

\begin{align}
  E &= \sum_{i=0}^{i=N-1} -\frac{h^3}{12}(v''(t_i))\\
  E &= -N\frac{h^3}{12}v''_{ave}\\
  E &= -\frac{t - t_0}{h} \frac{h^3}{12}v''_{ave}\\
  E &= -(t - t_0)\frac{h^2}{12}v''_{ave}\\
\end{align}

From this we can see that the Trapezoidal method's error is proportional to $h^2$.

\end{document}

>>>>

\section{3.2: Steady-state Population Model}

We will let $u_j < 1$ for all $j$. We also let the following be true:

\begin{equation*}
  \sum_j \beta_j \Big(\prod_{k<j}(1-u_k) \Big) = 1\\
\end{equation*}

Consider the following matrix:

\begin{equation}
  P = 
   \begin{pmatrix}
     \beta_1 & \beta_2 & ... & \beta_{N-1} & \beta_N\\
     1-u_1 & & & &\\
     & 1-u_2 & & &\\
     & & ... & &\\
     & & 1-u_{N-1} & &\\
     & & & 1-u_{N} & 0\\
   \end{pmatrix}
\end{equation}
\end{document}
